---
title: 'West'
output:
  distill::distill_article:
    toc:false:
      theme: theme.css
---

```{r required-packages, include = FALSE}
# For Analysis
library(readxl)
library(tidyverse)
library(doParallel)
library(R.utils)
library(gridExtra)
library(nlme)
library(viridis)


# For AQ_datadic and datadictionary
library(reactable)
library(htmltools)

# AQanalysis
library(lmtest)
library(ggplot2)

# For website
library(distill)

# For everyone
options(digits = 7)
registerDoParallel()
getDoParWorkers()
```


```{r reading_data, include = FALSE}
# census_detail
west_cd <- read_excel("data/Copy of nbhd_West Dallas_v9.xlsx", sheet = "census_detail")
# census_hh_inc
west_chi <- read_excel("data/Copy of nbhd_West Dallas_v9.xlsx", sheet = "census_hh_inc")
# land_use_by_year
west_lu <- read_excel("data/Copy of nbhd_West Dallas_v9.xlsx", sheet = "land_use_by_year")
# unit_counts_by_type
west_uc <- read_excel("data/Copy of nbhd_West Dallas_v9.xlsx", sheet = "unit_counts_by_type")
# median_sales_mls
west_msm <- read_excel("data/Copy of nbhd_West Dallas_v9.xlsx", sheet = "median_sales_mls")
```


## Executive Summary {-}

The report provides a longitudinal neighborhood and housing market analysis for the West area, spanning roughly 40 years for demographic trends and 20 years for land use and housing market dynamics. Overall, the West area shows:

* Significant racial and income composition shifts

* Declining population with rising housing costs

* Transition toward higher-value, owner-occupied housing

* Increasing housing market pressure and affordability challenges

These trends are consistent with gentrification and market tightening, with implications for housing stability and displacement risk.

Over the past four decades, the West area has undergone a pronounced demographic and housing transition marked by rising property values, declining renter presence, and a growing share of owner-occupied, single-family housing. Median home values, rents, and price per square foot have increased sharply since the mid-2010s, while total population and renter households have declined. These patterns indicate that housing cost escalation is outpacing resident retention.

## Population & Demographics {-}

This figure illustrates the long-term shift toward a majority-minority population, driven primarily by growth in the Hispanic population and a decline in the White population share.

```{r echo = FALSE, message=FALSE, warning=FALSE, fig.width=11.5,fig.height=5, layout="l-body-outset"}
west_cd$minority_pop <- west_cd$hispanic_pct + west_cd$black_pct
west_cd_vis <- west_cd %>%
  dplyr::select(white_pct, black_pct, minority_pop, hispanic_pct, asian_pct, amindian_pct, year...2) %>%
  gather(key = "variable", value = "value", -year...2) %>%
  mutate(variable = recode(variable, 'black_pct' = 'Black', 'white_pct' = 'White', 
                       'hispanic_pct' = 'Hispanic', 'asian_pct' = 'Asian', 'amindian_pct' = 'American Indian',
                       'minority_pop' = 'Minority Population'))


ggplot(west_cd_vis, aes(x = year...2, y = value, group=variable,colour=variable) ) + 
  geom_line(aes(color = variable), lwd = 1.5, linetype = "dashed") + 
  geom_point(lwd=3.5) +
  ylab("Percentage") +
  xlab("Year") + 
  scale_color_viridis(discrete = TRUE, option="plasma") +
  ggtitle("Population distribution by racial groups in the last 40 years")
```


## Socioeconomic Indicators {-}

Trends in median income, homeownership, poverty, household count, and population reveal rising housing costs alongside declining population and renter presence.

```{r echo = FALSE, message=FALSE, warning=FALSE, fig.width=13,fig.height=7, layout="l-body-outset"}
west_cd_vis <- west_cd %>%
  dplyr::select(total_households, owner_occupied, renter_occupied, poverty_rate, total_population, median_income_inflated_adj, year...2) %>%
  gather(key = "variable", value = "value", -year...2) %>%
  mutate(variable = recode(variable, 'poverty_rate' = 'Poverty Rate', 'total_population' = 'Total Population', 
                       'renter_occupied' = 'Renter Occupied', 'median_income_inflated_adj' = 'Median Income',
                       'total_households' = 'Total Households', 'owner_occupied' = 'Owner Occupied'))


ggplot(west_cd_vis, aes(x=year...2, y=value, group = 1))+
  geom_line(aes(color = variable), lwd = 1.5, linetype = "dashed")+
  geom_point(aes(color=variable), lwd=3.5) +
  facet_wrap(~variable, scales="free_y") +
  ylab("Value") +
  xlab("Year") + 
  scale_color_viridis(discrete = TRUE, option="turbo") +
  ggtitle("Demographic distribution in the last 40 years") +
  theme(legend.position="none")

```




```{r, include = FALSE}

west_cd_sub <- west_cd %>% dplyr::select(median_gross_rent_adj,year...2)


z_std <- function(observed) {
  #result <- (observed-observed[1])/(observed[1])*100
  result <- (observed)/sum(observed[1:4])*100
}

west_cd_sub <- west_cd %>% 
  select_if(is.numeric) %>% 
  mutate_all(funs(perc_ = z_std(.)))
```

## Household Size, Housing Value, and Tenure Trends {-}

Median home values and rents rise sharply over time, while renter-occupied households decline and owner occupancy increases.

```{r echo = FALSE, message=FALSE, warning=FALSE, fig.width=13,fig.height=7, layout="l-body-outset"}
west_cd$inflation <- c(0.124, 0.054, 0.034, 0.0316, 0.083)
west_cd$median_gross_rent_adj_inflation_adjusted <- west_cd$median_gross_rent_adj - west_cd$median_gross_rent_adj*west_cd$inflation
  
west_cd_vis <- west_cd %>%
  dplyr::select(renter_occupied, household_size, total_households, median_home_value_adj, median_gross_rent_adj, owner_occupied, year...2) %>%
  gather(key = "variable", value = "value", -year...2) %>%
  mutate(variable = recode(variable, 'household_size' = 'Household Size', 'total_households' = 'Total Households', 
                       'renter_occupied' = 'Renter Occupied', 'median_home_value_adj' = 'Median Home Value',
                       'median_gross_rent_adj' = 'Median Gross rent', 'owner_occupied' = 'Owner Occupied'))



ggplot(west_cd_vis, aes(x=year...2, y=value, group = 1))+
  geom_line(aes(color = variable), lwd = 1.5, linetype = "dashed")+
  geom_point(aes(color=variable), lwd=3.5)+
  facet_wrap(~variable, scales="free_y") +
  ylab("Value") +
  xlab("Year") + 
  scale_color_viridis(discrete = TRUE, option="turbo") +
  ggtitle("Demographic distribution in the last 40 years") +
  theme(legend.position="none")

```
                   
## Housing Cost Burden {-}

Cost-burdened and severely cost-burdened households persist, highlighting ongoing affordability challenges despite income growth.

```{r echo = FALSE, message=FALSE, warning=FALSE, fig.width=13,fig.height=7, layout="l-body-outset"}
west_cd_vis <- west_cd %>%
  dplyr::select(cost_burdened_units , cost_burdened_owner_units, severely_cost_burdened_units , severely_cost_burdened_owner_units, year...2) %>%
  gather(key = "variable", value = "value", -year...2) %>%
  mutate(variable = recode(variable, 'cost_burdened_units' = 'Cost Burdened Units',
                                     'cost_burdened_owner_units' = 'Cost Burdened Owner Units',
                                      'severely_cost_burdened_units' = 'Severely Cost Burdened Units',
                                       'severely_cost_burdened_owner_units' = 'Severely Cost Burdened Owner Units'))



ggplot(west_cd_vis, aes(x=year...2, y=value, group = 1))+
  geom_line(aes(color = variable), lwd = 1.5, linetype = "dashed")+
  geom_point(aes(color=variable), lwd=3.5)+
  facet_wrap(~variable, scales="free_y") +
  ylab("Value") +
  xlab("Year") + 
  scale_color_viridis(discrete = TRUE, option="turbo") +
  ggtitle("Demographic distribution in the last 40 years") +
  theme(legend.position="none")

```

## Household Income Distribution by Bracket {-}

Lower-income households decline substantially over time, while higher-income brackets become more prevalent, indicating economic filtering and displacement risk.

```{r include=FALSE}
west_chi_sub <- west_chi %>% replace(is.na(.), 0)
viz_func <- function(white, black, asian, hispanic, title){
  p <- ggplot(west_chi_sub, aes(x=`year`, group = 1))   + 
  geom_line(aes(y = white), color="red", lwd = 1.5, linetype = "dashed") +
  geom_point(aes(y = white), color="red", lwd=3.5)+
  geom_line(aes(y = black), color="blue", lwd = 1.5, linetype = "dashed") +
  geom_point(aes(y = black), color="blue", lwd=3.5)+
  geom_line(aes(y = asian), color="orange", lwd = 1.5, linetype = "dashed") +
  geom_point(aes(y = asian), color="orange", lwd=3.5)+  
  geom_line(aes(y = hispanic), color="purple", lwd = 1.5, linetype = "dashed") +
  geom_point(aes(y = hispanic), color="purple", lwd=3.5)+
  ylab("Count") +
  xlab("Year") + 
  scale_color_viridis(discrete = FALSE) +
  ggtitle(title) +
  scale_color_manual(name='Legends',
                     breaks=c('Black', 'White', 'Hispanic', 'Asian'),
                     values=c("White" = "red", "Black" = "blue", "Asian" = "orange", "Hispanic" = "purple")) + 
  theme(
    legend.position = c(0.5, 0.5),
    legend.justification = c("left", "top"), 
    legend.box.just = "left",
    legend.margin = margin(6, 6, 6, 6))
  return(p)
}
```



```{r echo = FALSE, message=FALSE, warning=FALSE, fig.width=20,fig.height=20, layout="l-body-outset"}
# Plot 1 - under $5000
p1 <- viz_func(west_chi_sub$income_under_5000_white, west_chi_sub$income_under_5000_black,
               west_chi_sub$income_under_5000_asian, west_chi_sub$income_under_5000_hispanic, "Income Under $5000") 

# Plot 2 - $5000 to $7500
p2 <- viz_func(west_chi_sub$income_5000_7500_white, west_chi_sub$income_5000_7500_black,
               west_chi_sub$income_5000_7500_asian, west_chi_sub$income_5000_7500_hispanic, "Income between $5000 to $7500") 
  
# Plot 3 - $7500 to $10000
p3 <- viz_func(west_chi_sub$income_7500_10000_white, west_chi_sub$income_7500_10000_black,
               west_chi_sub$income_7500_10000_asian, west_chi_sub$income_7500_10000_hispanic, "Income between $7500 to $10000") 
  
# Plot 4 - $10000 to $15000
p4 <- viz_func(west_chi_sub$income_10000_15000_white, west_chi_sub$income_10000_15000_black,
               west_chi_sub$income_10000_15000_asian, west_chi_sub$income_10000_15000_hispanic, "Income between $10000 to $15000") 

# Plot 5 - $15000 to $20000
p5 <- viz_func(west_chi_sub$income_15000_20000_white, west_chi_sub$income_15000_20000_black,
               west_chi_sub$income_15000_20000_asian, west_chi_sub$income_15000_20000_hispanic, "Income between $15000 to $20000") 

# Plot 6 - $20000 to $25000
p6 <- viz_func(west_chi_sub$income_20000_25000_white, west_chi_sub$income_20000_25000_black,
               west_chi_sub$income_20000_25000_asian, west_chi_sub$income_20000_25000_hispanic, "Income between $20000 to $25000") 

# Plot 7 - $25000 to $35000
p7 <- viz_func(west_chi_sub$income_25000_35000_white, west_chi_sub$income_25000_35000_black,
               west_chi_sub$income_25000_35000_asian, west_chi_sub$income_25000_35000_hispanic, "Income between $25000 to $35000") 

# Plot 8 - $35000 to $50000
p8 <- viz_func(west_chi_sub$income_35000_50000_white, west_chi_sub$income_35000_50000_black,
               west_chi_sub$income_35000_50000_asian, west_chi_sub$income_35000_50000_hispanic, "Income between $35000 to $50000") 

# Plot 9 - $25000 to $35000
p9 <- viz_func(west_chi_sub$income_50000_or_more_white, west_chi_sub$income_50000_or_more_black,
               west_chi_sub$income_50000_or_more_asian, west_chi_sub$income_50000_or_more_hispanic, "Income between $50000 or more") 


# income_35000_50000_white, income_50000_or_more_white
grid.arrange(p1, p2, p3, p4, p5, p6, p7, p8, p9, ncol=3, nrow=4)
```
## Land Use Distribution {-}

Single-family residential land use dominates development patterns, with limited growth in multifamily or mixed-density housing types.

```{r echo = FALSE, message=FALSE, warning=FALSE, fig.width=12,fig.height=7, layout="l-body-outset"}
west_lu_sub <- west_lu %>% dplyr::select(year, land_use, total_area_sqft)

west_lu_sub <- west_lu_sub %>% group_by(year) %>%
            summarize(Percentage = total_area_sqft / sum(total_area_sqft) * 100,
                      Land_Use = land_use)

ggplot(west_lu_sub, aes(x = year, y = Percentage, group=Land_Use,colour=Land_Use) ) + 
  geom_line(aes(color = Land_Use), lwd = 1.5, linetype = "dashed") + 
  geom_point(aes(color=Land_Use), lwd=3.5)+
  ylab("Percentage") +
  xlab("Year") + 
  scale_color_viridis(discrete = TRUE, option="plasma") +
  ggtitle("Land Use distribution in the last 20 years") + 
  labs(color='Land Use') 

```

## Housing Unit Composition by Type {-}

Growth in housing units is driven primarily by single-family homes, while duplexes and townhomes decline or stagnate.

```{r echo = FALSE, message=FALSE, warning=FALSE, fig.width=12,fig.height=10, layout="l-body-outset"}
west_uc_sub <- west_uc %>% dplyr::select(appraisal_yr, total_duplex_units, total_single_family_home_units, total_townhomes_units, total_apartment_units, total_all_units) %>%  replace(is.na(.), 0)


viz_func2 <- function(dataset, x_data, y_data, linecolor, title){
  p <- ggplot(dataset, aes(x = x_data, group = 1))   + 
  geom_line(aes(y = y_data), color=linecolor, lwd = 1.5, linetype = "dashed")+
  geom_point(aes(y = y_data), color=linecolor, lwd=3.5)+
  ylab("Units") +
  xlab("Year") + 
  scale_color_viridis(discrete = FALSE) +
  ggtitle(title)
  return(p)
}


p1 <- viz_func2(west_uc_sub, west_uc_sub$appraisal_yr, west_uc_sub$total_duplex_units, "red", "Total Duplex Units")
p2 <- viz_func2(west_uc_sub, west_uc_sub$appraisal_yr, west_uc_sub$total_single_family_home_units, "orange", "Total Single Family Home Units")
p3 <- viz_func2(west_uc_sub, west_uc_sub$appraisal_yr, west_uc_sub$total_townhomes_units, "purple", "Total Townhouse Units")
p4 <- viz_func2(west_uc_sub, west_uc_sub$appraisal_yr, west_uc_sub$total_apartment_units, "blue", "Total Apartment Units")
p5 <- viz_func2(west_uc_sub, west_uc_sub$appraisal_yr, west_uc_sub$total_all_units, "green", "Total Units")

grid.arrange(p1, p2, p3, p4, p5, ncol=2, nrow=3)
```


## Additional Demographic and Economic Trends {-}

Foreign-born population, senior share, income, rent, and poverty trends reflect an aging population and rising housing costs amid declining population.

```{r echo = FALSE, message=FALSE, warning=FALSE, fig.width=13,fig.height=7, layout="l-body-outset"}
west_cd_vis <- west_cd %>%
  dplyr::select(total_population, percent_seniors, percent_foreign_born, median_income_inflated_adj, poverty_rate, median_gross_rent_adj, year...2) %>%
  gather(key = "variable", value = "value", -year...2) %>%
  mutate(variable = recode(variable, 'total_population' = 'Total Population',
                          'percent_seniors' = 'Seniors %',
                          'percent_foreign_born' = 'Foreign Born %', 
                          'median_income_inflated_adj' = 'Median Income', 
                          'poverty_rate' = 'Poverty Rate',
                          'median_gross_rent_adj' = 'Median Rent'))


ggplot(west_cd_vis, aes(x=year...2, y=value, group = 1))+
  geom_line(aes(color = variable), lwd = 1.5, linetype = "dashed")+
  geom_point(aes(color=variable), lwd=3.5)+
  facet_wrap(~variable, scales="free_y") +
  ylab("Value") +
  xlab("Year") + 
  scale_color_viridis(discrete = TRUE, option="turbo") +
  ggtitle("Demographic distribution in the last 40 years") +
  theme(legend.position="none")

```
## Housing Market Dynamics {-}

Rising sale prices, price per square foot, and tight list-to-sale ratios demonstrate an increasingly competitive housing market with limited affordability.


```{r echo = FALSE, message=FALSE, warning=FALSE, fig.width=13,fig.height=15, layout="l-body-outset"}
west_cd_vis <- west_msm %>%
  dplyr::select(total_sales, median_close_price, median_ppsf, median_list_to_sale, list_to_sale_ratio_over_110, new_construction_count, median_year_built, sales_year) %>%
  gather(key = "variable", value = "value", -sales_year) %>%
  mutate(variable = recode(variable, 'total_sales' = 'Total Sales',
                          'median_close_price' = 'Median Close Price',
                          'median_ppsf' = 'Price Per Square Feet', 
                          'median_list_to_sale' = 'Median List to Sale', 
                          'list_to_sale_ratio_over_110' = 'List to Sale Ratio over 110',
                          'new_construction_count' = 'New Construction Count',
                          'median_year_built' = 'Median Year Built'))


ggplot(west_cd_vis, aes(x=sales_year, y=value, group = 1))+
  geom_line(aes(color = variable), lwd = 1.5, linetype = "dashed")+
  geom_point(aes(color=variable), lwd=3.5)+
  facet_wrap(~variable, scales="free_y", ncol = 2) +
  ylab("Value") +
  xlab("Year") + 
  scale_color_viridis(discrete = TRUE, option="turbo") +
  ggtitle("House Sale distribution in the last 20 years")+
  theme(legend.position="none") +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=1))
```





